A new tidy data structure to support exploration and modeling of temporal data

### 2019-08-20 tsibble.Rmd
done: intro revision

Temporal data arrives in many possible formats, with many different time contexts. For example, time can have various resolutions (hours, minutes, and seconds), and can be associated with different time zones with possible adjustments such as daylight saving time. Time can be regular (such as quarterly economic data or daily weather data), or irregular (such as patient visits to a doctor's office). Temporal data also often contains rich information: multiple observational units of different time lengths, multiple and heterogeneous measured variables, and multiple grouping factors. Temporal data may comprise the occurrence of events, such as flight departures, that need to be reduced to a regular structure.

Despite this variety and heterogeneity of temporal data, current software typically requires time series objects to be model-oriented matrices. Analysts are expected to do their own data preprocessing and take care of anything else needed to allow model fitting, which leads to a myriad of ad hoc solutions and duplicated efforts.

@r4ds proposed the tidy data workflow, which provides a conceptual framework for processing data (as described in Figure \@ref(fig:tsibble-fit)). Currently, time series modeling and forecasting enters this framework at the *modeling* stage, while temporal data enters at the start. This paper integrates time series analysis into this tidy framework, providing a coherent way for getting temporal data into the matrix format for modeling. These objects were designed for modeling methods. The diagram in the style of Figure \@ref(fig:tsibble-fit) would place the model at the center of the analytical universe, and all the transformations and visualizations would hinge on that format. This is contrary to the **tidyverse** conceptualization, which holistically captures the full data workflow.

@@

(ref:tsibble-fit-cap) Illustration of the data science workflow, drawn from @r4ds, showing how current time series tools interface with the workflow and how the tsibble structure and tools integrate. The new data structure, tsibble, makes the connection between temporal data input, and specialist modeling formats. It provides elements at the "tidy" step, which produce tidy temporal data for time series visualization and modeling.

@@

Time series can be univariate or multivariate, and for modeling require relatively long lengths (i.e., large $T$). Time series researchers and analysts who are concerned with this large $T$ property, are mostly concerned with stochastic processes, for the primary purpose of forecasting, and characterizing temporal dynamics. Most statistical software represent such time series as vectors or matrices. Multivariate time series are typically assumed to be in the format where each row is assumed to hold observations at a time point and each column to contain a single time series. (The tidy data name for this would be **wider format**.) This implies that data are columns of homogeneous types: numeric or non-numeric, but there are limited supporting methods for non-numeric vectors. Time indices are implicitly inferred as attributes or meta-information. The number of observations must be the same across all the series. Data wrangling, from the form that data arrives in to this specialist format, can be mysterious and frustrating, inhibiting the robustness of downstream tasks.

@@

Stata [@Stata] provides two commands, `xtset` and `tsset`, to declare time series and panel data respectively, both of which need specifying panel and time variables. Note they are in fact the same command, according to the manual. The underlying data arrangement is the same and of longer form. Either time series or panel commands are sufficient to apply to both types. The SAS software [@SAS-STAT] handles them in a similar way as Stata.

@@

Most time series models are univariate, such as ARIMA and Exponential Smoothing, modelling temporal dependency for each series independently. The **fable** package [@R-fable] provides a tidy forecasting framework built on top of tsibble is under development, with the goal of promoting transparent and human-centered forecasting practices. With the presence of the key, tsibble holds many series. Since models are fundamentally scalable, the `model()` and `forecast()` generics will take care of fitting and forecasting univariate models to each series across time in a tsibble at once.

@@

>>>>

Temporal data arrives in many possible formats, with many different time contexts. For example, time can have various resolutions (hours, minutes, and seconds), and can be associated with different time zones with possible adjustments such as daylight saving time. Time can be regular (such as quarterly economic data or daily weather data), or irregular (such as patient visits to a doctor's office). Temporal data also often contains rich information: multiple observational units of different time lengths, multiple and heterogeneous measured variables, and multiple grouping factors. Temporal data may comprise the occurrence of time-stamped events, such as flight departures.

Despite this variety and heterogeneity of temporal data, the lack of a uniform data organization expects analysts to do their own data preprocessing and take care of anything else needed to allow further analysis, which leads to a myriad of ad hoc solutions and duplicated efforts.

@r4ds proposed the tidy data workflow, which provides a conceptual framework for exploring data (as described in Figure \@ref(fig:tsibble-fit)). In temporal domain, data with temporal information arrives at the "import" stage. A new abstraction, *tsibble*, is the gatekeeper at the "tidy" stage, to verify if the raw temporal data is appropriate for downstream analytics. The exploration loop will be aided with declarative grammars for more robust and accurate analysis.

@@

(ref:tsibble-fit-cap) Illustration of the data science workflow regarding temporal data, drawn from @r4ds. The new data structure, tsibble, makes the connection between temporal data input, and downstream analytics. It provides elements at the "tidy" step, which produce tidy temporal data for temporal visualization and modeling.

@@

Time series can be univariate or multivariate, and for modeling require relatively long lengths (i.e., large $T$). Time series researchers and analysts who are concerned with this large $T$ property, are mostly concerned with stochastic processes, for the primary purpose of forecasting, and characterizing temporal dynamics. Most statistical software represent such time series as vectors or matrices. Multivariate time series are typically assumed to be in the format where each row is assumed to hold observations at a time point and each column to contain a single time series. (The tidy data name for this would be **wider format**.) This implies that data are columns of homogeneous types: numeric or non-numeric, but there are limited supporting methods for non-numeric vectors. Time indices are implicitly inferred as attributes or meta-information. The number of observations must be the same across all the series. Data wrangling, from the form that data arrives in to this specialist format, can be mysterious and frustrating, inhibiting the robustness of downstream tasks. This kind of format is model-centric, and the diagram in the style of Figure \@ref(fig:tsibble-fit) would place the "model" at the center of the analytical universe, and all the transformations and visualizations would hinge on that format. This is contrary to the **tidyverse** conceptualization, which holistically captures the full data workflow.

@@

Stata [@Stata] provides two commands, `tsset` and `xtset`, to declare time series and panel data respectively, both of which need specifying panel and time variables. Note they are in fact the same command, according to the manual. The underlying data arrangement is the same and of longer form. Either time series or panel commands are sufficient to apply to both types. The SAS software [@SAS-STAT] handles them in a similar way as Stata.

@@

Most time series models are univariate, such as ARIMA and Exponential Smoothing, modelling temporal dynamics for each series independently. The **fable** package [@R-fable] provides a tidy forecasting framework built on top of tsibble is under development, with the goal of promoting transparent and human-centered forecasting practices. With the presence of the key, tsibble holds many series. Since models are fundamentally scalable, the `model()` and `forecast()` generics will take care of fitting and forecasting univariate models to each series across time in a tsibble at once.

@@

* extensive index
* flexible key
